---
title: "Notes and Code from Summer Quarter (all classes)"
author: "Clarissa Boyajian"
date: "12/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

# EDS 212: Essential Math for Environmental Data Science

### Logs and exponents in R
Some examples of using the `log()` function and `exp()` **function** in R
```{r}
# Examples of logarithms
log(4)

# Example of exponents
exp(5)

# A test of log()
log(exp(1))
```

### Creating sequences of values
```{r}
# A sequence from 1 to 20, by increments of 1
ex_1_212.1 <- seq(from = 1, to = 20, by = 1)

# Vector from 0 to 40 by increments of 0.2
ex_2_212.1 <- seq(from = 0, to = 40, by = 0.2)

# Vector from 1 to 5000 by increments of 10
ex_3_212.1 <- seq(from = 1, to = 5000, by = 10)
```

### Create a function for logistic growth
Given the carrying capacity, initial population, growth rate, and time, use the logistic growth model to estimate the population at that time.
```{r}
pop_logistic <- function(capacity, init_pop, rate, time_yr) {
  capacity / (1 + ((capacity - init_pop) / init_pop) * exp(-rate * time_yr))
}
  
pop_logistic(capacity = 2580, init_pop = 230, rate = 0.32, time_yr = 2.4)
```

### Try finding population over a sequence of times
```{r}
# Create a vector of times from 0 to 20 by increments of 0.5
time_vec <- seq(from = 0, to = 45, by = 0.25)
```

Now let's estimate the population over that range of times, given capacity = 2580, initial population = 230, rate = 0.32
```{r}
chipmunk_pop <- pop_logistic(capacity = 2580, init_pop = 230, rate = 0.32, time_yr = time_vec)
```

### Create a data frame and plot chipmunk population over time
```{r}
chipmunk_df <- data.frame(time_vec, chipmunk_pop)
```

### Create a bare minimum ggplot graph:
```{r}
ggplot(data = chipmunk_df, aes(x = time_vec, y = chipmunk_pop)) +
  geom_point() + 
  labs(x = "Time (years)", 
       y = "Chipmunk Population (individuals)",
       title = "Inyo National Forest Chipmunks",
       subtitle = "Yay Chipmunks")
```

### Creating functions
```{r}
G <- function(t, z) {
  3.1 * ((t-4.2) ^ 2) + 0.06*z
}

G(t = 1, z = 2.5)
```

```{r}
P <- function(initial, rate, time) {
  initial*exp(rate*time)
}

P(initial = 2500, rate = 0.046, time = 4)
```

### End of Day Practice
Make function to calculate fish weight from a, L, and b, calculate weights for one species, create table (sequence) of lengths and weights, plot lenght vs. weight
```{r 212 day 1}
# Create fish weight function
W <- function(a, L, b) {
  a * L^b}

# Milkfish weights
W(a = 0.0905, b = 2.52, L = 60)

# Giant trevally weights
W(a = 0.0353, b = 3.05, L = 60)

# Great barracuda weights
W(a = 0.0181, b = 3.27, L = 60)

# Create df sequence of size ranges from 0-200 by 1 cm
fish_length <- seq(from = 0, to = 200, by = 1)

barracuda_weight <- W(a = 0.0181, b = 3.27, L = fish_length)

# Bind lengths and estimate barracuda lengths in one df
barrracuda_df <- data.frame(fish_length, barracuda_weight)

# Create ggplot graph of lenghts vs. weights
ggplot(data = barrracuda_df, aes(x = fish_length, y = barracuda_weight)) + 
  geom_point() +
  labs(x = "Barracuda length (cm)",
       y = "Barracuda weight (g)",
       title = "Comparing Great Barracudas")
```








# EDS 221: Scientific Programming Essentials










